### 2012-11-06 paper/isca40format/Introduction.tex
stuff reasonable

\section{Introduction} With the increasing number of cores, bus-based
cache-coherence protocols have started being replaced by message-passing
directory-based protocols. Bus-based protocols allow only one cache-line
transaction to happen at a time, locking the bus till the transaction is
complete. This serializes memory accesses across each processor, even if they
are for different cache lines. On the other hand, message-passing,
directory-based protocols can handle multiple independent transactions in
parallel, thus scaling better with increasing core counts.

In a message-passing directory-based protocol, the coherence state changes are
orchestrated by passing messages in the form of upgrade or downgrade requests
and responses between the caches via the directory. Unlike bus-based protocols
which lock the bus till all the required state transitions are done thereby
ensuring atomicity of state transitions, in a message-passing protocol, requests
and responses are split leading to transient states. This creates scenarios not
present in bus-based protocols, which can lead to coherence violations or
deadlocks, if the protocol is not designed or implemented carefully.

For instance, consider a scenario in which a cache having read permission for
a line gets a store request from its core and hence requires write permissions
for that line. The cache would send an upgrade request to the directory
controller and may eventually get a response from the directory controller. The
response would signify that the permissions in the other caches \emph{were}
compatible with this cache having write permission for the line, \ie the other
caches neither \emph{had} read nor write permissions for that line. But the messages
take time to propagate. Thus, as emphasized, the response message is based on
the old permissions that the other caches had, not the permissions that the
other caches have at the moment the original cache gets the response. If an
other cache gets a read or write permission for that line during the time the
messages are in flight, coherence would be violated.

Consider a different scenario in which a cache, already having read permission
for a line, gets a store request, and hence sends an upgrade request for write
permission to the directory controller. Instead of getting a response, the cache
may receive a request from the directory controller asking this cache to give up
the read permission for the line \ie an invalidate request. This could happen if
the directory controller received a request for write permission from another
cache earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its request
for write permission, or would it deadlock? As we will be showing later, this
could create a deadlock.

Consider yet another scenario in which a cache receives an invalidate request
from the directory controller for a line that is not present in the cache. Is
such a request even possible? The directory does not have access to the
up-to-date state of the cache, and so can potentially make this request. How
should a cache handle this request; should it drop such a request, or respond
saying that it does not have the line? Again, reasoning about these questions is
very difficult, as one has to take into account all the possible transitions
that can happen in the entire system.

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing directory-based setting.
Even for a 2-level cache hierarchy, a private L1 cache for each core and a
shared L2 cache, designing and implementing the protocol is very hard. This
problem is greatly exacerbated in a multi-level cache hierarchy, which is
becoming the norm because of the increasing core counts.

In this paper, we present a systematic design methodology to design hierarchical
messaging-passing, directory-based inclusive cache coherence protocols. We
restrict our methodology to inclusive protocols as we want our protocols to
scale to multiple levels of hierarchy. A non-inclusive or exclusive cache
coherence protocol requires probing every level of the hierarchy starting from
the root (LLC) to the leaf (L1 cache) to process any request for permission. We
believe that this would not scale as the number of cores and the number of
cache levels increase.

We present a set of transitions that the caches and directories must follow on
sending or receiving a request or a response. We have formally proved that
these behaviors will ensure the following properties: a) the directory's notion
of a cache's state is ``conservative'' in that a cache is never going to have
more permissions that what the directory thinks the cache has, and b) any
request will eventually get a corresponding response. It should be noted that
the formal proof was written using the Coq theorem prover \cite{}, and the
discussion of the formal proof is beyond the scope of this paper.

The flexibility of our methodology to design multiple cache coherence protocols
comes from the fact that we do not assign any meaning or semantics to the value
of a coherence state. All the transitions that we present are in the form of
upgrades or downgrades from state $x$ to state $y$, without giving concrete
values for $x$ or $y$. We can thus use our methodology to design multiple
hierarchical cache coherence protocols: MSI, MESI, MOSI, MOESI. We will show
how each of these protocols obey the set of transitions that we require for our
methodology, proving the correctness of all these protocols.

%\paragraph{Paper organization:} Section \ref{sec:setup} gives the setup for our
%methodology. It gives the state transitions of the cache and directory
%controllers on sending or receiving messages. Section \ref{sec:properties} gives
%the two properties that are guaranteed if the state transitions that we present
%are obeyed. We present the intuition for proving these properties starting from
%the state transitions that we present in \ref{sec:setup}. In Section
%\ref{sec:protocols}, we show how these properties can be used to implement a
%hierarchical MSI, MESI, MOSI and MOESI protocols. We also demonstrate how the
%correctness and deadlock freedom of each of the protocols follows straightaway
%from our methodology with no additional verification effort. Our claim is that
%our state transitions and the guaranteed properties can be used to formally
%prove any message-passing directory-based hierarchical inclusive cache coherence
%protocols by appropriate mapping. In Section \ref{sec:performance}, we show that
%these formally verified family of protocols require no extra state bits, nor
%extra message transfers compared to other published protocols. In Section
%\ref{sec:related}, we present some of the related work and finally in Section
%\ref{sec:conclusion}, we present the directions we would like to take this work
%in the future and conclude.

%We carried out the proof using the Coq theorem prover \cite{}. Unlike
%model-checking based techniques, the theorem prover merely checks the
%correctness of a supplied mathematical proof, thereby avoiding the state
%explosion problem. The proof is not restricted to a fixed number of cores,
%number of caches levels or number of caches in each level. Instead the proof can
%be parameterized for these quantities, thus ensuring that the same proof holds
%for any configuration, without any additional computational complexity. The
%downside of a theorem prover is that it shifts the onus of supplying a correct
%mathematical proof to the user, instead of searching through the state space to
%expose violations of specifications \ala model checking.

%$S$ and $I$ in an MSI protocol
%The state transitions that we require does not enforce a particular cache
%coherence protocol. It merely dictates how the states change on sending or
%receiving a response message, or after handling a request message. To be more
%concrete, consider a simple example. A directory controller receives a Write
%permission request for a line from a cache. If the directory thinks that some
%other cache has a read or write permission for that line, then a correct cache
%coherence protocol requires that the other cache invalidates that line before
%the directory sends a response to the first cache giving it Write permission.
%But this ``policy'' is not enforced by the state transitions that we present.
%Instead, if the directory decides to give Write permission to the original cache
%when it thinks another cache still has that line, then our state transitions
%will permit it. The system will no longer be coherent, but it will not violate
%the two properties that our state transitions guarantee (the directory would now
%think that the original cache has Write permission and some other caches also
%have some permissions for that line). This enables us to implement a wide range
%of correct cache coherence protocols with minimal proof burden.
%
%For example, it is straight forward to implement MSI or MESI protocols making
%use of the properties guaranteed in our methodology. If the directory controller
%receives a Write permission request from a cache for a line, then it sends
%requests to every cache which has that line. Our second property of getting a
%response for every request ensures that these caches would respond eventually,
%making their local states for that line Invalid. Our first property guarantees
%that when the directory controller thinks the state of a cache for a line is
%Invalid, then the actual states in the caches at that time should not be greater
%than Invalid. Thus, once all the other caches have responded, the directory
%controller can respond to the original cache. This way, the two properties can
%be used to implement MSI or MESI in a straightforward manner. We will present
%how this is done in greater detail in the paper.
%
%A more complicated protocol with cache intervention (like MOSI and MOESI)
%can also be implemented using the two properties that we guarantee, but it
%requires an additional step of mapping the protocol to our transitions (and
%proving that the mapping is correct). We will present this in greater detail in
%the paper.
%
%Finally, we will show that these formally verified family of protocols require
%no extra state bits, nor extra message transfers compared to other published
%protocols. In fact, our claim is that our state transitions and the two
%guaranteed properties can be used to formally prove any message-passing
%directory-based hierarchical inclusive cache coherence protocols by
%appropriately mapping the protocol to our transitions, as shown in the
%MOSI/MOESI mapping.

>>>>

\section{Introduction}

Large shared-memory multicores are likely to use cache-coherence protocols
which are directory based, hierarchical, and implemented using message passing.
Even though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can not
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

In this paper we will consider systems where caches form a tree hierarchy and
caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
parent is guaranteed to contain address $a$, though the parent may have stale
data for $a$. Every cache or node in the system has an associated directory
which contains information about its children caches. The coherence state
changes are orchestrated by passing messages in the form of \emph{upgrade} or
\emph{downgrade} requests and responses between the caches via the directory.
Unlike bus-based protocols which lock the bus till all the required state
transitions are done thereby ensuring atomicity of state transitions, in a
message-passing protocol, requests and responses are split leading to transient
states. This creates scenarios not present in bus-based protocols, which can
lead to coherence violations or deadlocks, if the protocol is not designed or
implemented carefully.

Several factors contribute to the difficulty of designing protocols for such
distributed, hierarchical and message-passing systems:

\begin{enumerate}

\item Lots of concurrent events must be dealt with without a central authority
to impose a sequence on events;

\item The state on which transitions are based is physically distributed in
space and cannot be accessed atomically;

\item Physical network may block the transmission of a message because of lack
of communication buffers or head-of-the-line blocking;

\item For each message type, determining which cache and directory states are
possible and therefore must be dealt with is hard. An incomplete set of
transition rules may cause the protocol to deadlock.

\end{enumerate}

As examples of problems in designing correct protocols, we describe several
concrete and familiar scenarios from two-level systems (a private L1 cache for
each processor and a shared L2 cache):

\noindent \emph{Example 1:} A cache with a read permission for a line gets a
store request from its processor and hence, requires a write permission for
that line. The cache would send an upgrade request to the directory controller
and may eventually get a response from the directory controller. The response
would signify that the states in the other caches \emph{were} compatible with
this cache having write permission for the line. That is, the other caches
neither \emph{had} read nor write permissions for that line. But the messages
take time to propagate and the response message is necessarily based on the old
state bits (permissions) that the other caches had, not the permissions that
the other caches have at the moment the original cache gets the response. If
another cache gets a read or write permission for that line during the time the
messages are in flight, coherence would be violated.

\emph{Example 2:} A cache, already having read permission for a line, gets a
store request, and sends an upgrade request for write permission to the
directory controller. Instead of getting a response, the cache may receive a
request from the directory controller asking this cache to give up the read
permission for the line \ie an invalidate request. This could happen if the
directory controller received a request for write permission from another cache
earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its
request for write permission, or would it deadlock?

\emph{Example 3:} A cache receives an invalidate request from the directory
controller for a line that is not present in the cache. Is such a request even
possible? The directory does not have access to the up-to-date state of the
cache, and so can potentially make this request. How should a cache handle this
request; should it drop such a request, or respond saying that it does not have
the line?

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing directory-based setting.
Even for a 2-level cache hierarchy, designing and implementing the protocol is
very hard. This problem is greatly exacerbated in a multi-level cache
hierarchy, which is becoming the norm because of the increasing core counts.

In this paper, we present a substantially easier method to design
cache-coherence protocols for such systems. The method essentially breaks down
the design problem in two parts: a part that is an abstract protocol shared by
a number of concrete protocols, and another part that is specific to the
protocol to be designed and implemented. The abstract protocol is parameterized
by the number of cache levels and the number of caches in each level of the
hierarchy as well as by a predicate which involves a local ``compatibility of
state'' test. We illustrate our methodology by designing multiple
hierarchical cache-coherence protocols, notably MSI, MESI, MOSI, MOESI, both in
the presence as well as in the absence of point-to-point FIFO networks.

Our method substantially reduces the design and proof burden for new protocols
because of the common abstract protocol. The abstract protocol is a set of
transitions that the caches and directories follow to send or receive a request
or a response. We have formally proved that these transitions ensure the
following properties: a) the directory's notion of a cache's state is
``conservative'' in that a cache is never going to have more permissions that
what the directory thinks the cache has; and b) any request will eventually get
a corresponding response. A formal proof of the abstract protocol was written
using the Coq theorem prover \cite{}, however, a discussion of the formal proof
is beyond the scope of this paper.

It should be noted that the flexibility of our methodology to design multiple
cache coherence protocols comes from the fact that we do not assign any meaning
to the value of a coherence state beyond upgrades or downgrades. Thus, how or
why a particular cache state implies that a particular address may not be
present in any other cache stems from protocol specific ``compatibility of
state'' check.

The main contributions of this paper are: 1. A new abstract protocol that
satisfies two important invariants for distributed and hierarchal
message-passing protocols; 2. Instantiation of the abstract protocol to
implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their
variations and showing why their correctness is guaranteed as long as the
abstract protocol is correct, \ie it provably implements the two invariants;
and 3. Our method does not introduce any overhead over direct implementation of
these protocols.  We think that this new method of designing protocols
substantially reduces the required effort, and that our abstract protocol is
flexible enough to implement (while ensuring correctness of) most if not all
inclusive message-passing directory-based protocols.

\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract
protocol; Section \ref{sec:properties} gives the two properties that are
guaranteed by the abstract protocol and provides some intuition for proving
these properties. We also discuss the communication buffer requirements for the
abstract protocol and show that the requirement is only proportional to the
number of levels and not to the number of processors. In Section
\ref{sec:protocols}, we show how the abstract protocol can be used to implement
MSI, MESI, MOSI and MOESI protocols and their variants. In Section
\ref{sec:performance}, we show that these formally verified family of protocols
require no extra state bits, nor extra message transfers compared to other
published implementations. In Section \ref{sec:related}, we present some of the
related work and finally in Section \ref{sec:conclusion}, we present some
directions for the future work.
